---
title: "covariate_f_with_error"
author: "Emma Govan"
date: "2023-04-17"
output: html_document
---


$$y_{ij} \sim N\left(\frac{\sum_{k=1}^Kp_{ik}q_{k}(\mu_c+\mu_s)}{\sum_{k=1}^Kp_{ik}q_{k}}, \frac{\sum_{k=1}^Kp_{ik}^2q_{k}^2(\sigma_c^2+\sigma_s^2)}{\sum_{k=1}^Kp_{ik}^2q_k^2} +\tau^{-1}\right)$$



$$\pi(y\mid\theta) = \frac{1}{\sqrt{2\pi}}\left(\frac{\sum_{k=1}^Kp_{ik}^2q_{k}^2(\sigma_c^2+\sigma_s^2)}{\sum_{k=1}^Kp_{ik}^2q_k^2} +\tau^{-1}\right)^{\frac{n}{2}}\exp\left(\frac{\frac{\sum_{k=1}^Kp_{ik}^2q_{k}^2(\sigma_c^2+\sigma_s^2)}{\sum_{k=1}^Kp_{ik}^2q_k^2} +\tau^{-1}}{2}\sum_{i=1}^n(y_i-\frac{\frac{\sum_{k=1}^Kp_{ik}q_k(\mu_s+\mu_c)}{\sum_{k=1}^Kp_{ik}q_k}}{2})^2\right)$$



$$log(\pi(y\mid\theta) = -\frac{1}{2}\log(2\pi) +\frac{n}{2}\log\left(\frac{\sum_{k=1}^Kp_{ik}^2q_{k}^2(\sigma_c^2+\sigma_s^2)}{\sum_{k=1}^Kp_{ik}^2q_k^2} +\tau^{-1}\right)- \frac{\frac{\sum_{k=1}^Kp_{ik}^2q_{k}^2(\sigma_c^2+\sigma_s^2)}{\sum_{k=1}^Kp_{ik}^2q_k^2} +\tau^{-1}}{2}\left(\sum_{i=1}^N\left(y_i - \frac{\sum_{k=1}^Kp_{ik}q_k(\mu_s+\mu_c)}{\sum_{k=1}^Kp_{ik}q_k}\right)^2\right)$$




$$p_{ik} = \frac{\exp{f_{ik}}}{\sum_{k=1}^K\exp{f_{ik}}}$$

$$f \sim mvn(\alpha_k +\beta_kx_{ik}, \Sigma_f)$$

$$\pi(f) = \frac{1}{(2\pi)^\frac{n}{2}}\frac{1}{\mid\Sigma_f\mid^\frac{1}{2}} \exp\left(-\frac{1}{2}(f-\alpha_k-\beta_kx_{ik})^T\Sigma^{-1}_k(f-\alpha_k-\beta_kx_{ik})\right)$$

$$\log(\pi(f)) = -\frac{n}{2}\log(2\pi) -\frac{1}{2}\log(\mid\Sigma_f\mid) -\frac{1}{2}(f-\alpha_k-\beta_kx_{ik})^T\Sigma^{-1}_k(f-\alpha_k-\beta_kx_{ik})$$

$$\pi(\tau) = \frac{d^c}{\Gamma(c)} \tau_j^{c-1} \exp(-d\tau_j)$$

$$\log(\pi(\tau)) = c\log(d) -log(\Gamma(c)) - (c-1)\log(\tau_j) -d\tau_j)$$

All we need to specify for FFVB is h and log_q

$$h(\theta) = \log(\pi(p)\pi(\tau)\pi(y|p, \tau))$$

This has $$c_0, d_0, \Sigma_{f0}$$ etc


$$\log{q_\lambda}(\theta) = \log(q(f)q(\tau))$$

Which uses the new iterative value of $\mu_f, \Sigma_f, c$, and $d$ generated in every iteration of the model

sim_theta will also have to be specified, this generates new values of $\tau$ and $f$ every iteration

Everything else is already done in the generic simmr functions.

$\theta = p, \tau$

$\lambda = \alpha_f, \beta_f, \Sigma_f,  c, d$



# Read in data
```{r}
library(tidyverse)
library(readxl)
library(R2jags)

path = system.file("extdata", "geese_data.xls", package = "simmr")
geese_data = lapply(excel_sheets(path), read_excel, path = path)
sex <- geese_data[[1]]$Sex[1:9]
consumer <- read.csv("geese_consumer.csv", header = TRUE, stringsAsFactors = FALSE)
sources <- read.csv("geese_sources.csv", header = TRUE, stringsAsFactors = FALSE)[1:4,]
disc <- read.csv("geese_discrimination.csv", header = TRUE, stringsAsFactors = FALSE)
y<-(consumer %>% filter(Group == "1") %>% select(d15N, d13C))[1:9,]
```







The hyper-parameter values for the prior distributions:
```{r}

n_isotopes <- 2

mu_kj <- sources[,c(2,4)] #+ disc[, c(2,4)]
K<-nrow(mu_kj)
sigma_kj <- sources[,c(3,5)]
mu_c <- disc[, c(2,4)]
sigma_c <- disc[, c(3,5)]
q <- sources[, c(6:7)]

# sigma_kj<-matrix(rep(0, K*n_isotopes), nrow = K, ncol = n_isotopes)
# mu_c<-matrix(rep(0, K*n_isotopes), nrow = K, ncol = n_isotopes)
# sigma_c<-matrix(rep(0, K*n_isotopes), nrow = K, ncol = n_isotopes)
#q<-matrix(rep(1/K, K*n_isotopes), nrow = K, ncol = n_isotopes)

```


#JAGS
JAGS CODE
```{r}


model_code <- "
model{
  for (i in 1:N) {
    for (j in 1:J) { 
      y[i,j] ~ dnorm(inprod(p[i,]*q[,j], s_mean[,j]+c_mean[,j]) / inprod(p[i,],q[,j]), 1/var_y[i,j])
      var_y[i,j] <- inprod(pow(p[i,]*q[,j],2),s_sd[,j]^2+c_sd[,j]^2)/pow(inprod(p[i,],q[,j]),2) 
        + pow(sigma[j],2)
    }
  }
  for(i in 1:N) {
    p[i,1:K] <- expf[i,]/sum(expf[i,])
    for(k in 1:K) {
      expf[i,k] <- exp(f[i,k])
      f[i,k] = mu_f[i,k]
    }
  }
  for(k in 1:K) { 
  for(i in 1:N) {
      mu_f[i,k] <- alpha[k] +beta[k]*x[i]
      }
      
    alpha[k] ~ dnorm(0,1)
    beta[k] ~dnorm(0,1) 
  }

  
  for(j in 1:J) { sigma[j] ~ dgamma(0.001, 0.001) }
}
"

model_data = list(y=y,
                 s_mean=mu_kj,
                 s_sd=sigma_kj,
                 c_mean=mu_c,
                 c_sd=sigma_c,
                 q=q, 
                 N=nrow(y),
                 K=length(sources$Sources),
                 J=ncol(y),
                 x = c(rep(1,9))
                 )
            
model_parameters <- c("alpha", "beta", "sigma", "p")

# Run the model
model_run <- jags(
  data = model_data,
  parameters.to.save = model_parameters,
  model.file = textConnection(model_code),
  n.chains = 4, # Number of different starting positions
  n.iter = 10000, # Number of iterations
  n.burnin = 2000, # Number of iterations to remove at start
  n.thin = 5
) # Amount of thinning)


print(model_run)
```



##FFVB

```{r}
# Source in all the generic functions
source("FF_VB_generic_functions_correct.R")
## Pretty sure these functions should work fine, don't need to be changed - everything thats
## put into them should be a sum so just a single number anyway

# Set up data and priors --------------------------------------------------
# Hyper-parameters - vague priors
S<-100
x_i0 <- sex
n <- nrow(y)
mu_alpha_zero <- c(rep(0, K))
mu_beta_zero <- c(rep(0, K))
sigma_alpha_zero<-c(rep(1,K))
sigma_beta_zero<-c(rep(1,K))
n_isotopes <- ncol(mu_c)
c_0 <- c(rep(1, n_isotopes))
d_0 <- c(rep(1, n_isotopes))

#TEMP
sex<-c(rep(1,9))



# Functions for this model ------------------------------------------------

## This is working when beta is set to zero so think its okay
sim_theta <- function(S, lambda) {
  
  # For K parameters you will have
  # lambda is of length K+K*(K+1)/2 +n_isotopes*2
  # mean <- lambda[1:K]
  # chol_prec is made up of lambda[(K + 1):(K+(K*(K+1))/2)]
  # Tau is made up of lambda[((K+(K*(K+1))/2)+1):((K+(K*(K+1))/2)+n_isotopes*2)]
  # (f) ~ MVN(lambda[1:K], solve(crossprod(chol_prec)))
  
  mean_alpha <- lambda[1:K]
 
  # K*(K-1) precision terms
  chol_prec_alpha <- matrix(0, nrow = K, ncol = K)
  chol_prec_alpha[upper.tri(chol_prec_alpha, diag = TRUE)] <- lambda[(K + 1):(K + (K * (K + 1)) / 2)]
 
  mean_beta <- lambda[(K+(K*(K+1)/2) +1):(2*K + (K*(K+1)/2))]
  chol_prec_beta <- matrix(0, nrow = K, ncol = K)
  chol_prec_beta[upper.tri(chol_prec_beta, diag = TRUE)] <- lambda[(2*K + (K*(K+1)/2) + 1):(K*(K+1) +2*K)]
  
  # This is a placeholder for more advanced code using chol_prec directly
  theta <- cbind(
    t(rMVNormC(S, mu = mean_alpha, U = chol_prec_alpha)),
    t(rMVNormC(S, mu = mean_beta, U = chol_prec_beta)), #matrix(rep(0.1, K*S), ncol = K, nrow = S), ##
    matrix(rgamma(S * n_isotopes,
                  shape = lambda[((K*(K+1) +2*K)+1):((K*(K+1) +2*K)+n_isotopes)],
                  rate = lambda[((K*(K+1) +2*K)+n_isotopes+1):((K*(K+1) +2*K)+2*n_isotopes)]
    ),
    nrow = S,
    ncol = n_isotopes,
    byrow = TRUE
    )
  )

  return(theta)
}

#theta <- sim_theta(S, lambda)
#Theta is alpha, beta, and tau
#lambda is mu_alpha, sigma_alpha, mu_beta, sigma_beta, c, d


# Log of likelihood added to prior
h <- function(theta) {
  # Create alpha and beta
  alpha<- theta[1:K]
  beta<-theta[(K+1):(2*K)]
  sigma<-theta[(2*K + 1):(2*K + 2)]
  f<-matrix(NA, ncol = K, nrow = n)
  for(i in 1:n){
    for(k in 1:K){
  f[i,k]<-alpha[k] + beta[k] * sex[i]
    }
  }
  p<-matrix(NA, ncol = K, nrow = n)
  for(i in 1:n){
  p[i,] <- exp(f[i,1:K]) / (sum((exp(f[i,1:K]))))
  }
  
  ##Im not sure this bit needs to be looped over?
  hold<-0

    for(i in 1:n){
   hold = hold + sum(dnorm(y[i, 1],
                   mean = sum(p[i,] * q * (mu_kj[, 1]+mu_c[,1]))/sum(p[i,]*q),
                   sd = sqrt(sum(p[i,]^2 * q^2 * (sigma_kj[, 1]^2+sigma_c[,1]^2))/sum(p[i,]^2*q^2) + 1 / theta[2 *K + 1]),
                   log = TRUE
  ))
     sum(dnorm(y[i, 2],
                   mean = sum(p[i,] * q[,2] * (mu_kj[, 2]+mu_c[,2]))/sum(p[i,]*q[,2]),
                   sd = sqrt(sum(p[i,]^2 * q[,2]^2 * (sigma_kj[, 2]^2+sigma_c[,2]^2))/sum(p[i,]^2*q[,2]^2) + 1 / theta[2 *K + 2]),
                   log = TRUE
  ))
    } 
  
  
  return(hold + sum(dnorm(alpha, mu_alpha_zero, sigma_alpha_zero, log = TRUE)) +
    sum(dnorm(beta, mu_beta_zero, sigma_beta_zero, log = TRUE)) +
    sum(dgamma(sigma, shape = c_0, rate = d_0, log = TRUE)) )
}
# h(theta[1,])

log_q <- function(lambda, theta) {
  
  mean_alpha <- lambda[1:K]
  chol_prec_alpha <- matrix(0, nrow = K, ncol = K)
  chol_prec_alpha[upper.tri(chol_prec_alpha, diag = TRUE)] <- lambda[(K + 1):(K + (K * (K + 1)) / 2)]
 
  mean_beta <- lambda[(K+(K*(K+1)/2) +1):(2*K + (K*(K+1)/2))]
  chol_prec_beta <- matrix(0, nrow = K, ncol = K)
  chol_prec_beta[upper.tri(chol_prec_beta, diag = TRUE)] <- lambda[(2*K + (K*(K+1)/2) + 1):(K*(K+1) +2*K)]  
  
  
  # This is a placeholder for more advanced code using chol_prec directly
  # prec <- crossprod(chol_prec)
  p1 <- matrix(theta[1:K] - mean_alpha, nrow = 1) %*% t(chol_prec_alpha)
  p2<-matrix(theta[(K+1):(2*K)] - mean_beta, nrow = 1) %*% t(chol_prec_beta)
  # log_det <- unlist(determinant(prec, logarithm = TRUE))["modulus"]
  return(-0.5 * K * log(2 * pi) - 0.5 * sum(log(diag(chol_prec_alpha))) - 0.5 * p1%*%t(p1) -
          0.5 * K * log(2 * pi) - 0.5 * sum(log(diag(chol_prec_beta))) - 0.5 * p2%*%t(p2)
         
         + sum(dgamma(theta[(2*K+1):(2*K+2)],
                      shape = lambda[((K*(K+1) +2*K)+1):((K*(K+1) +2*K)+n_isotopes)],
                      rate = lambda[((K*(K+1) +2*K)+n_isotopes+1):((K*(K+1) +2*K)+2*n_isotopes)],
                      log = TRUE
         )))
}
#log_q(lambda, theta[1,])

# Algorithm ---------------------------------------------------------------
lambda <- run_VB(lambda = c(rep(0,K), rep(1,K*(K+1)/2), rep(0,K), rep(1,K*(K+1)/2), rep(1,n_isotopes), rep(1,n_isotopes))) # Starting value of lambda


```
#Compare to JAGS
```{r}
mycol <- rgb(0, 0, 255, max = 255, alpha = 125, names = "blue50")
mycol2 <- rgb(255, 0, 0, max = 255, alpha = 125)

theta<-sim_theta(6400, lambda)

#  f<-matrix(NA, ncol = K, nrow = n)
# 
# for(i in 1:n){
#     for(k in 1:K){
#   f[i,k]<-theta[i,k] + theta[i,(k+K)] * sex[i]
#     }
#   }
# 
# p<-matrix(NA, ncol = K, nrow = n)
#   for(i in 1:n){
#   p[i,] <- exp(f[i,1:K]) / (sum((exp(f[i,1:K]))))
#   }


# Plot of alpha and beta
hist((model_run$BUGSoutput$sims.list$alpha[,1]), col = mycol2, breaks = 80)
hist(theta[,1], col = mycol, alpha = 0.5, breaks = 10, add = TRUE)

hist((model_run$BUGSoutput$sims.list$alpha[,2]), col = mycol2, breaks = 80)
hist(theta[,2], col = mycol, alpha = 0.5, breaks = 10, add = TRUE)

hist((model_run$BUGSoutput$sims.list$alpha[,3]), col = mycol2, breaks = 80)
hist(theta[,3], col = mycol, alpha = 0.5, breaks = 10, add = TRUE)

hist((model_run$BUGSoutput$sims.list$alpha[,4]), col = mycol2, breaks = 80)
hist(theta[,4], col = mycol, alpha = 0.5, breaks = 10, add = TRUE)

hist((model_run$BUGSoutput$sims.list$beta[,1]), col = mycol2, breaks = 80)
hist(theta[,5], col = mycol, alpha = 0.5, breaks = 10, add = TRUE)

hist((model_run$BUGSoutput$sims.list$beta[,2]), col = mycol2, breaks = 80)
hist(theta[,6], col = mycol, alpha = 0.5, breaks = 10, add = TRUE)

hist((model_run$BUGSoutput$sims.list$beta[,3]), col = mycol2, breaks = 80)
hist(theta[,7], col = mycol, alpha = 0.5, breaks = 10, add = TRUE)

hist((model_run$BUGSoutput$sims.list$beta[,4]), col = mycol2, breaks = 80)
hist(theta[,8], col = mycol, alpha = 0.5, breaks = 10, add = TRUE)


#Plots of sigma
hist((model_run$BUGSoutput$sims.list$sigma[,1]), col = mycol2, breaks = 80)
hist(1/sqrt(theta[,9]), col = mycol, alpha = 0.5, breaks = 80, add = TRUE)

hist((model_run$BUGSoutput$sims.list$sigma[,2]), col = mycol2, breaks = 80)
hist(1/sqrt(theta[,10]), col = mycol, alpha = 0.5, breaks = 80, add = TRUE)

  alpha<- colMeans(theta[,1:K])
  beta<-colMeans(theta[,(K+1):(2*K)])
  sigma<-colMeans(theta[,(2*K + 1):(2*K + 2)])
  f<-matrix(NA, ncol = K, nrow = n)
  for(i in 1:n){
    for(k in 1:K){
  f[i,k]<-alpha[k] + beta[k] * sex[i]
    }
  }
  p<-matrix(NA, ncol = K, nrow = n)
  for(i in 1:n){
  p[i,] <- exp(f[i,1:K]) / (sum((exp(f[i,1:K]))))
  }




```



```{r}
library(simmr)

simmr_in<-simmr_load(mixtures = y, 
                     source_names = sources$Sources, 
                     source_means = mu_kj, 
                     sigma_kj, 
                     mu_c, 
                     sigma_c, 
                     q)

simmr_out<-simmr_mcmc(simmr_in)
summary(simmr_out)


##JAGS code is right - have chekced by setting beta to zero
##Need to go into 
```








